---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You can install DEUS from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("timjeske/DEUS")
```

### Prerequisites

- [CD-Hit](http://weizhongli-lab.org/cd-hit/) (cd-hit-est binary)
- [BLAST](ftp://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/) (blastn binary)
- Custom BLAST database used for annotation step


### Preparing a BLAST database
During the annotation step, a BLAST database is used to determine possible biological sources of each unique sequence. Depending on the use-case, two options are available how an appropriate database can be obtained:

1. Use existing databases available at e.g. [NCBI](ftp://ftp.ncbi.nlm.nih.gov/blast/db/). Database sets may also be retrieved automatically with update_blastdb.pl, which is part of the BLAST+ suite.
2. Build a custom database using BLAST makeblastdb and some fasta sequences:

      cat DASHR.fa Homo_sapiens.GRCh37.75.cds.all.fa  Homo_sapiens.GRCh37.75.ncrna.fa > HumanDB.fa
      makeblastdb -in HumanDB.fa -dbtype nucl -parse_seqids
